---
title: "Application on MOCA 100k"
author: "Alexis Vandenbon & Diego Diez"
date: "2020-09-17"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Application on MOCA 100k}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
library(here)
library(tidyverse)
library(Seurat)
library(singleCellHaystack)

knitr::opts_chunk$set(
	fig.align = "center",
	fig.width = 7,
	message = FALSE,
	warning = TRUE
)
```

# Load data

Here we apply `haystack` to 100k cells from the Mouse Organogenesis Cell Atlas (MOCA). The sparse matrix data was downloaded from the [MOCA website](https://oncoscape.v3.sttrcancer.org/atlas.gs.washington.edu.mouse.rna/landing). The data was converted into a Seurat object and processed following the [standard pipeline](https://satijalab.org/seurat/v3.2/pbmc3k_tutorial.html).

```{r eval=FALSE}
library(here)
library(tidyverse)
library(Seurat)
library(singleCellHaystack)
library(scmisc)
```


```{r load}
x <- read_rds(here("data-raw/moca_100k.rds"))
x
```


```{r plot_dim}
DimPlot(x, label = TRUE) + NoLegend() + NoAxes()
```
 
 
# Haystack

We run `haystack` using the method _highD_ with the default threshold for detection, which is more than 1.


```{r eval=FALSE}
system.time({
  res <- haystack(x, method = "highD")
})
```


```{r haystack, echo=FALSE}
r <- system.time({
  res <- haystack(x, method = "highD")
})
r
```


It takes `r format(unname(r)[3] / 60, digits = 0)` minutes to complete in a standard personal computer. Here we show the top 10 genes selected by `haystack`.

```{r top}
top <- show_result_haystack(res, n = 10)
top
```

And here we plot the expression of the top 4 genes.

```{r plot_results}
FeaturePlot(x, rownames(top)[1:4], order = TRUE) & NoAxes()
```


