---
title: "Application on toy example"
author: "Alexis Vandenbon"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Application on toy example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5,
  fig.align = 'center'
)

dat.tsne <- singleCellHaystack::dat.tsne
dat.expression <- singleCellHaystack::dat.expression
```

# Application on a toy dataset

A small toy dataset is included in the package. The toy dataset includes:

- `dat.expression`: scRNA-seq expression of genes (rows) in cells (columns)

- `dat.tsne`:       a 2D representation of the cell in dat.expression


First, apply `haystack` (the main function of the package) on the toy dataset. This should take just several seconds on a typical desktop computer.

```{r example1}
library(singleCellHaystack)
set.seed(1234)

# Turn the expression data into detection (gene detected = TRUE, not detected = FALSE)
dat.detection <- dat.expression > 1

# run the main 'haystack' analysis
res <- haystack(dat.tsne, detection = dat.detection, method = "2D")

# the returned results 'res' is of class 'haystack'
class(res)
```

Let's have a look at the "most surprising" genes.

```{r example2}
# show top 10 "surprising" genes
show_result_haystack(res.haystack = res, n=10)

# alternatively: use a p-value threshold
#show_result_haystack(res.haystack = res, p.value.threshold = 1e-10)
```

One of the most non-random genes is "gene_497". Here we visualize it's expression in the t-SNE plot.

```{r example3}
# visualize one of the surprizing genes
plot_gene_haystack(
  dat.tsne,
  expression = dat.expression,
  gene = "gene_497",
  detection = dat.detection,
  high.resolution = TRUE,
  point.size = 2
)
```

Yes, the pattern of cells in the toy example roughly resembles a haystack; see [the Haystack paintings by Monet](https://en.wikipedia.org/wiki/Haystacks_(Monet_series)).

# Clustering and visualization

You are not limited to single genes. Here, we pick up a set of non-random genes, and group them by their expression pattern in the plot into 5 clusters.

```{r example4}
# get the top most significant genes, and cluster them by their distribution pattern in the 2D plot
sorted.table <- show_result_haystack(res.haystack = res, p.value.threshold = 1e-10)
gene.subset <- row.names(sorted.table)

# k-means clustering
km <- kmeans_haystack(dat.tsne, detection=dat.detection, genes=gene.subset, k=5)
km.clusters <- km$cluster

# alternatively: hierarchical clustering
#hc <- hclust_haystack(dat.tsne, detection=dat.detection, genes=gene.subset)
#hc.clusters <- cutree(hc,k = 5)
```

... and visualize the average pattern of the genes in cluster 1 (for example).

```{r example5}
# visualize cluster distributions
plot_gene_set_haystack(dat.tsne, detection = dat.detection, 
                        genes=names(km.clusters[km.clusters==1]), point.size=2)
```

