---
title: "Clustering singleCellHaystack results into gene modules"
author: "Diego Diez"
date: "2022-10-09"
output: html_notebook
---

```{r setup}
library(ggplot2)
library(Seurat)
library(SeuratData)
library(singleCellHaystack)
library(ComplexHeatmap)

theme_set(cowplot::theme_cowplot())

select_cell_markers <- function(scores, max.overlap=1) {
  sel <- apply(scores, 1, function(x) {
    pos <- x > 0
    npos <- sum(pos)
    if (npos > 0 & npos <= max.overlap)
      return(TRUE)
    return(FALSE)
  })
  #rownames(scores)[sel]
  scores[sel,]
}

set.seed(1)
```

# Introduction

In this vignette we use the PBMC dataset from 10x available in the `SeuratData` package.

```{r}
pbmc <- LoadData("pbmc3k", "pbmc3k.final")
pbmc
```

```{r}
DimPlot(pbmc, label=TRUE) + NoLegend() + NoAxes()
```

# Run haystack

```{r}
coord <- Embeddings(pbmc, reduction="pca")
dim(coord)
```

```{r}
exprs <- GetAssayData(pbmc, assay="RNA", slot="data")
exprs[1:4, 1:4]
```

```{r}
res <- haystack_continuous_highD(coord, exprs)
```

```{r}
plot_rand_fit(res, "mean")
plot_rand_fit(res, "sd")
```

```{r}
sum <- show_result_haystack(res)
head(sum)
```

```{r}
ggplot(sum, aes(x=seq_len(nrow(sum)), y=log.p.adj)) +
  geom_point() + 
  geom_vline(xintercept=c(100, 1000), color=c("limegreen", "violetred"))
```

Next we look at the top 100 genes identified by haystack.

```{r}
top_100_haystack <- head(sum, 100)
head(top_100_haystack)
```

# Using whole expression profiles

We can use the top genes to calculate gene clusters using the single cell expression data.

```{r}
m <- as.matrix(exprs[rownames(top_100_haystack), ])
m[1:3, 1:3]
```

```{r fig.width=4, fig.height=8}
Heatmap(m, column_split=pbmc$seurat_annotations, name="Expression", show_column_names=FALSE, column_title_rot=45)
```

# Usingn cluster-wise averages

We can calculate cluster scores for each gene. Basically we compute the average of each gene in each cluster.

```{r}
scores <- AverageExpression(pbmc, features=rownames(top_100_haystack), assays="RNA")[["RNA"]]
head(scores)
```
```{r}
genes.pass <- apply(scores, 1, function(x) {
  sum(x > 1)
})
table(genes.pass)
```

```{r}
Heatmap(t(scale(t(scores[genes.pass, ]))))
```

We need to scale the scores.

```{r}
scores <- t(scale(t(scores)))
head(scores)
```
We can see that there are clear cluster specific markers.

```{r fig.width=4, fig.height=8}
Heatmap(scores, name="Score")
```
# Using the distribution P

Another approach is to use the distribution P used by haystack. This is the density weighted expression at the grid points. Haystack returns by default the density.

```{r}
scores <- res$info$P
```

